A Linear Solution of the Linearity Challenge

### 2025-12-26 paper/conclusion.tex
global revision

We have presented a formalisation of \LCC, a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} supported by the same type system
of \CP~\cite{Wadler14}. Binary sessions are modeled in \LCC using the
continuation-passing encoding described by \citet[extended
version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
%
Under the assumption that the size of a formalisation is a good measure of its
complexity, the formalisation of \LCC can be qualified as ``simple'' compared to
the existing ones.
%
In this respect, our work casts some doubts on the actual contribution of
context splitting to the complexity of a formalisation and hints instead at
explicit continuations in the underlying model as a simplifying factor.
%
It is quite striking that, among the approaches that have been proposed to
overcome context splitting, the one of \citet{SanoKavanaghPientka23} makes use
of explicit continuations.

We remark that \LCC and the calculus described in the linearity
challenge~\cite{CarboneEtAl24} are incomparable in terms of expressiveness. On
the one hand, the challenge only considers a minimal calculus of first-order,
monomorphic sessions while \LCC is a fully-fledged calculus that supports
linear, shared, higher-order, polymorphic channels. On the other hand, the
logical foundations of \LCC prevent the modeling of deadlocked processes and of
sequential processes that own both endpoints of a session. These processes can
be modeled in the calculus of the challenge, where parallel composition and name
restriction are distinct constructs as in the $\pi$-calculus.
%
Because of its logical foundations, we like to think of \LCC as a ``canonical''
calculus that deserves its own space in the context of the linearity challenge
alongside with (but not in substitution of) more traditional session calculi.

The simple formalisation of \LCC is a good basis for further investigations. We
have already extended \LCC with support for infinite types and recursive
processes (the source code of this extension is publicly available in the same
repository as the one that contains the formalisation of \LCC). In the future,
it would be interesting to formalise the strong normalisation property of \LCC
as a consequence of cut elimination of classical linear logic.

In this work we have focused on models and theories of \emph{binary sessions}
(those connecting exactly two processes), but there are also formalisations of
\emph{multiparty sessions}, notably those by \citet{JacobsBalzerKrebbers22a} and
\citet{TiroreBengstonCarbone25}, which can be significantly more complex than
those of binary sessions. The formalisation by \citet{JacobsBalzerKrebbers22a}
amounts to 173kb and the one by \citet{TiroreBengstonCarbone25} to more than 1Mb
of Coq code. Also in these cases, the formalisation based on (virtual)
continuations~\citep{JacobsBalzerKrebbers22a} happens to be substantially
smaller. Whether this is a coincidence or further evidence of the effectiveness
of the continuation-based approaches is left for future work.

>>>>

We have presented a formalisation of \LCC, a linear calculus of continuations
closely related to the linear $\pi$-calculus~\cite{KobayashiPierceTurner99} and
supported by the same type system of \CP~\cite{Wadler14}. Binary sessions can be
modeled in \LCC using the continuation-passing encoding described by
\citet[extended version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
%
Under the assumption that the size of a formalisation is a good measure of its
complexity, the formalisation of \LCC qualifies as the simplest compared to the
existing ones. This result should also be weighed considering the richness of
\LCC in terms of supported features and provable properties.

Our work casts some doubts on the actual contribution of context splitting to
the complexity of a formalisation and hints instead at explicit continuations in
the underlying model as a simplifying factor.
%
It is quite remarkable that, among the alternative approaches that have been
proposed to avoid context splitting, the one of \citet{SanoKavanaghPientka23}
based on linearity predicates makes use of explicit continuations.

The calculus of linear continuations and the calculus described in the linearity
challenge~\cite{CarboneEtAl24} are incomparable in terms of expressiveness. On
the one hand, the challenge only considers a minimal calculus of first-order,
monomorphic sessions while \LCC is a fully-fledged calculus that supports
linear, shared, higher-order, polymorphic channels. On the other hand, the
logical foundations of \LCC prevent the modeling of deadlocked processes and of
sequential processes that own both endpoints of a session. These processes can
be modeled in the calculus of the challenge, where parallel composition and name
restriction are distinct constructs as in the $\pi$-calculus.
%
Because of its logical foundations, we think that \LCC deserves its own space in
the context of the linearity challenge alongside with (but not in substitution
of) more traditional session calculi.

The simple formalisation of \LCC is a good starting point for further
developments. We have already extended \LCC with support for infinite types and
recursive processes (the source code of this extension is publicly available in
the same repository as the one that contains the formalisation of \LCC). In the
future, it would be interesting to formalise the strong normalisation property
of \LCC as a consequence of cut elimination of classical linear logic.

In this work we have focused on models and theories of \emph{binary sessions}
(those connecting exactly two processes), but there are also formalisations of
\emph{multiparty sessions}, notably those by \citet{JacobsBalzerKrebbers22a} and
\citet{TiroreBengstonCarbone25}, which can be significantly more complex than
those of binary sessions. The formalisation by \citet{JacobsBalzerKrebbers22a}
amounts to 173kb and the one by \citet{TiroreBengstonCarbone25} to more than 1Mb
of Coq code. Also in these cases, the formalisation based on (virtual)
continuations~\citep{JacobsBalzerKrebbers22a} happens to be substantially
smaller. Whether this is a coincidence or further evidence of the effectiveness
of the continuation-based approaches is left as future work.

### 2025-12-26 paper/related.tex
global revision

Motivated by the technical difficulties arising from context splits,
\citet{SanoKavanaghPientka23} define a structural version of \CP using an
alternative approach based on \emph{linearity predicates}. The key idea is to
treat typing contexts structurally and to enforce the linear usage of channels
by checking their syntactical occurrence in processes. Interestingly, the
approach relies on the \emph{explicit naming of continuations} so as to
precisely account for the number of times a channel is actually used.
\citet{SanoKavanaghPientka23} do not connect their technique with the
continuation-passing encoding of binary sessions, but \emph{a posteriori} the
analogies between the two approaches are evident.

\citet{ZackonEtAl25} describe yet another approach to the handling of typing
contexts whereby channels are associated not just with a type but also with a
\emph{tag} (or \emph{multiplicity}) from a given resource algebra. The tag
summarises the number of allowed usages of a particular channel, including the
possibility that the channel is not available. This approach streamlines context
splitting since contexts can be treated in an essentially structural way, except
for tags which are conveninently combined using operations from the resource
algebra.

@@

\Cref{tab:comparison} reports a summary of the aforementioned formalisations and
also our own. The three leftmost colums list the reference papers (sorted by
publication date), the proof assistant and the formalised calculus.
%
For simplicity, we have grouped the calculi into classes identified by an
acronym. Some acronyms, like \CP~\cite{Wadler14} and
\GV~\cite{GayVasconcelos10,Wadler14} are well known in the literature on session
types. It should be noted that the actual calculus being formalised usually
differs from (typically, is a strict subset of) the one identified by the
acronym, but it retains at least some of its key features. We use \SessionPI to
denote (variants of) the session-based $\pi$-calculus presented by
\citet{HondaVasconcelosKubo98} and \LinearPI to denote (variants of) the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99}, into which binary sessions can be
encoded using explicit continuation channels as described by \citet[extended
version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}. Finally, we use
\SCP for the structural version of \CP introduced by
\citet{SanoKavanaghPientka23} and \Calculus for our calculus, a linear version
of \CP with explicit continuations.

@@

The rightmost columns of \Cref{tab:comparison} report the size (in kilobytes) of
the formalisations as rough (and possibly questionable) estimates of their
complexity. Instead of counting line numbers, which may be affected by code
indentation styles and syntactical constraints of the proof assistant being
used, we have counted the number of characters after comments have been removed
and spaces have been squeezed.\footnote{Sequences of two or more consecutive
space-like characters are collapsed into a single space. The squeezing is
obtained by running the command \texttt{tr -s [:space:] file} on Unix-like
systems.}
%
The reported sizes account for the source code of the formalisation excluding
examples and any safety result, if present. We have noticed that the
interpretation of ``safety'' varies widely among the formalisations and, except
for our own, they all differ from the one stated in the linearity
challenge~\citep{CarboneEtAl24} when available.
%
Some formalisation define \emph{libraries} which could be reused in different
contexts. In these cases, the size of the library is reported separately next to
the size of the part of the development that uses it.

@@

Based on the available data, we can make the following observations.
%
The adoption of context splitting, which is very well represented, does not seem
to be a good indicator of the complexity of the formalisation. Indeed, the
formalisations based on context splitting span the whole range of sizes, from
the largest by \citet{GotoEtAl16} (543kb) to our own (21kb, without the proof of
deadlock freedom) which is also the only one supporting all the featurs of \CP.

@@

We found no strong evidence that the intrinsically typed representation of
processes reduces the size of the formalisation. As observed in
\Cref{sec:process-agda}, this choice may help reducing the overall number of
datatypes to be defined, but the definitions are also more involved because they
incorporate invariants and the proofs of typing preservation. In the end, it
seems plausible to speculate that the effort for representing processes, types
and typing rules is not substantially different overall.

>>>>

Motivated by the technical difficulties arising from context splittings,
\citet{SanoKavanaghPientka23} define a structural version of \CP using an
approach based on \emph{linearity predicates}. The key idea is to treat typing
contexts structurally and to enforce the linear usage of channels by checking
their syntactical occurrence in processes. Interestingly, this approach relies
on the \emph{explicit naming of continuations} so as to precisely account for
the number of times a channel is actually used. \citet{SanoKavanaghPientka23} do
not connect their technique with the continuation-passing encoding of binary
sessions, but the analogies are evident even though the role of continuations
differs.

\citet{ZackonEtAl25} describe yet another approach for managing typing contexts
whereby channels are associated not just with a type but also with a \emph{tag}
(or \emph{multiplicity}) from a given resource algebra. The tag summarises the
number of allowed usages of a particular channel, including the possibility that
the channel is not available. This approach streamlines context splitting since
contexts can be treated in an essentially structural way, except for tags which
are conveninently combined using operations from the resource algebra.

@@

\Cref{tab:comparison} reports a summary of the aforementioned formalisations and
also our own. The three leftmost columns list the reference papers (sorted by
publication date), the used proof assistant and the formalised calculus.
%
For simplicity, we have grouped the calculi into classes identified by an
acronym. The acronyms \CP~\cite{Wadler14} and
\GV~\cite{GayVasconcelos10,Wadler14} refer to well-known calculi in the
literature on session types. It should be noted that the actual calculus being
formalised usually differs from (typically, is a strict subset of) the one
identified by the acronym, but it retains at least some of its key features.
%
The acronym \SessionPI refers to (variants of) the session-based $\pi$-calculus
presented by \citet{HondaVasconcelosKubo98} while \LinearPI refers to (variants
of) the linear $\pi$-calculus~\cite{KobayashiPierceTurner99}, into which binary
sessions can be encoded using explicit continuation channels as described by
\citet[extended version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
Finally, \SCP is the structural version of \CP introduced by
\citet{SanoKavanaghPientka23} and \Calculus is our calculus.

@@

The rightmost columns of \Cref{tab:comparison} report the size (in kilobytes) of
the formalisations as rough (and possibly questionable) estimates of their
complexity. Instead of counting line numbers, which may be affected by code
indentation styles and syntactical constraints of the proof assistant being
used, we have counted the number of characters after comments have been removed
and spaces have been squeezed.\footnote{Sequences of two or more consecutive
space-like characters are collapsed into a single space. The squeezing is
obtained by running the command \texttt{tr -s [:space:] file} on Unix-like
systems.}
%
The reported sizes account for the source code of the formalisation excluding
examples and any safety result, if present. We have noticed that the
interpretation of ``safety'' varies widely among the formalisations and, except
for our own, they all differ from the one stated in the linearity
challenge~\citep{CarboneEtAl24}.
%
Some formalisations define \emph{libraries} which could be reused in different
contexts. In these cases, the size of the library is reported separately next to
the size of the part of the development that uses it.

@@

Based on the available data, we can make the following observations.
%
The adoption of context splitting, which is very well represented, does not seem
to be a good indicator of the complexity of the formalisation. Indeed, the
formalisations based on context splitting span the whole range of sizes, from
the largest by \citet{GotoEtAl16} (543kb) to our own (21kb, without the proof of
deadlock freedom) which is also the only one supporting all the features of \CP.

@@

We found no strong evidence that the intrinsically typed representation of
processes reduces the size of the formalisation. As observed in
\Cref{sec:process-agda}, this choice may help reducing the overall number of
datatypes to be defined, but the definitions are also more involved because they
incorporate invariants and the proofs of typing preservation. It seems plausible
to speculate that the effort for representing processes, types and typing rules
is not substantially different overall.
